---
editor: 
  markdown: 
    wrap: sentence
---

```{r}
# data manipulation
library(tidyverse)
# birthwt data
library(MASS)
# draw dag
library(dagitty)
```

# Paradox I

Suppose you collect data on intelligence and diligence among students.
Further suppose that aptitude is a function of intelligence and diligence.
We define students with high aptitude as "good students."

```{r}
# data for Berkson's paradox
set.seed(2023)  
N <- 1000
intelligence <- rnorm(N)
diligence. <-  rnorm(N)
aptitude <- 1/2 * intelligence + 1/2 * diligence. + rnorm(N, 0, .1)

df_1 <- data.frame(intelligence = intelligence,
                         diligence. = diligence.,
                         aptitude = aptitude) %>% 
  mutate(good_students = aptitude > 0)
```

1.  Among good students, visualize and describe the relationship between intelligence and diligence.

```{r}
good_students <- filter(df_1, good_students)
ggplot(good_students, aes(x=intelligence, y=diligence.)) +
  geom_point() +
  geom_smooth(method=lm, color= "orange") +
  labs(title="Intelligence vs Diligence among Good Students", subtitle="There is a somewhat linear and negative relationship between intelligence and 
diligence among good students", x="Intelligence", y="Diligence")

```

2.  Using the whole population, visualize and describe the relationship between intelligence and diligence.

```{r}
ggplot(df_1, aes(x=intelligence, y=diligence.)) +
  geom_point() +
  geom_smooth(method=lm) + 
  labs(title="Intelligence vs Diligence among all students", subtitle="There is no pattern between intelligence and diligence among the entire student population ", x="Intelligence", y="Diligence")
```

3.  How do you resolve this paradox? Use visualization.

```{r}
ggplot(df_1, aes(x=intelligence, y=diligence., colour = good_students)) +
  geom_point() +
  geom_smooth(method=lm) + 
  labs(title="Intelligence vs Diligence among all students", subtitle="", x="Intelligence", y="Diligence", colour= "Good Students")
```

4.  Draw a DAG to explain this paradox.

```{r}
dag {
Aptitude
[outcome,pos="-0.382,-0.395"]
Diligence
[selected,pos="0.437,-0.133"]
Intelligence 
[selected,pos="-1.107,-0.135"]
Diligence -> Aptitude 
Intelligence -> Aptitude
}

```

# Paradox II

```{r}
data(birthwt)
glimpse(birthwt)
```

1.  Calculate the proportion of low birth weight (LBW) infant by mother's smoking status. Is this what you expected?

```{r}
```

2.  Visually investigate whether age plays an important role. What do you observe?

```{r}
```

3.  Draw a DAG explaining this phenomena. Help: https://cran.r-project.org/web/packages/dagitty/vignettes/dagitty4semusers.html

```{r}
# modify this
g1 <- dagitty("dag {
    W1 -> Z1 -> X -> Y
    Z1 <- V -> Z2
    W2 -> Z2 -> Y
    X <-> W1 <-> W2 <-> Y
}")
plot(graphLayout(g1))
```

# Paradox III

Students are assigned to a diet program.
To evaluate the program, you measure their weights before (pre-intervention) and after the program (post-intervention).

Suppose Analyst A uses the change in weight as the response variable.
On the other hand, Analyst B uses the post-intervention weight as the response variable but conditions on (or controls/adjusts for) the pre-intervention weight.

```{r data}
set.seed(2023)
N <- 1000
 
b <- 10
l <- 50
u <- 70
Male <- runif(N, l, u)
Female <- Male + b
 
b1 <- 0.5
b0_M <- (1 - b1) * mean(Male)
b0_F <- mean(Female) - b1 * (mean(Male) + b)
 
sd <- 1
Male1 <- b0_M + b1 * Male + rnorm(N, sd=sd)
Female1 <- b0_F + b1 * Female + rnorm(N, sd=sd)

df_2 <- data.frame(pre = c(Male, Female), post = c(Male1, Female1))
df_2$sex <- c(rep("Male", N), rep("Female", N))
```

1.  Visualize Analyst 1's approach. What is their conclusion?

```{r}
```

2.  Visualize Analyst 2's approach. What is their conclusion?

```{r}
```

3.  Draw a DAG to explain their approaches.

```{r}
```
